---
title: "Forecasting Daily PM2.5 Levels in San Diego"
author: "Eva Chow, Sai Thiha, and Andrew Abeles"
date: "11/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(plotly)
library(astsa)
library(tseries)
library(vars)
library(fastDummies)
library(forecast)
```

# Data Preparation and Exploration

```{r}
# load dataset
airquality <- read.csv('SD_airquality_2016-2020.csv')
```


```{r}
# data summary
summary(airquality)
``` 

```{r}
# data visualizations (convert Date into date data type first)
airquality$Date <- as.Date(airquality$Date, format="%m/%d/%Y")

# looking at PM2.5 across all five years
firstfigure <- airquality %>%
  ggplot( aes(x=Date, y=PM2.5.AQI.Value)) +
  geom_area(fill='darkblue', alpha=0.5) +
  geom_line(color='darkblue') +
  ylab('PM 2.5 AQI Value') + 
  ggtitle('PM 2.5 AQI Data from 2016-2020') +
  theme(plot.title=element_text(size=14, face="bold"), 
        axis.title.y=element_text(size=12, face="bold"),
        axis.title.x=element_text(size=12, face="bold"))

firstfigure <- ggplotly(firstfigure)
firstfigure
```

OBSERVATIONS:
- quick observations show that many peak PM2.5 readings occur in Q4 of the year
- interestingly, some of the lowest readings occur around MAR/APR
- sharp decrease in PM2.5 readings in MAR 2020 (possibly related to COVID lockdowns?)
- max PM2.5 reading from SEP-OCT 2020 (coinciding with fires)

```{r}
# add additional columns for year and day of the year
airquality <- transform(airquality,
                        Year = format(Date, '%Y'),
                        DayOfYear = as.numeric(format(Date, '%j')))
```

```{r}
#year comparisons of PM2.5
secondfigure <- airquality %>%
  ggplot( aes(x=DayOfYear, y=PM2.5.AQI.Value, group=Year, colour=Year)) +
  geom_point() + 
  geom_line() +
  ylab('PM 2.5 AQI Value') + 
  ggtitle('PM 2.5 AQI Data - Year Comparisons') +
  theme(plot.title=element_text(size=14, face="bold"), 
        axis.title.y=element_text(size=12, face="bold"),
        axis.title.x=element_text(size=12, face="bold"))

secondfigure <- ggplotly(secondfigure)
secondfigure
```

OBSERVATIONS:
- 2016 has erratic readings during Q1, including both highest and lowest PM2.5 of that year
- 2017 has peak readings during last two months of year
- 2018 has peaks around JUL and from DEC-FEB
- 2019 peaks around OCT, but has noticeably less variation in spikes compared to previous years
- 2020 has the biggest range, with the lowest of the 5 year readings occurring MAR2020 and the highest occuring SEP2020
- 2020 has a general trend upwards as year progresses (random walk?)
majority of PM2.5 readings occur between 50-100

Let's create a time series object for the daily PM2.5 data. 
```{r}
PM2.5 <- ts(
  airquality$PM2.5.AQI.Value,
  start = c(2016, 1), # the series starts on the 1st day of 2016
  frequency = 365.25 # number of days per year, accounting for leap years 
)
PM2.5
```

## Stationarity 

```{r}
# Augmented Dickey-Fuller test 
adf.test(PM2.5)
```

The p-value is 0.01, so we can reject the null hypothesis that the time series has a unit root in favor of the alternative hypothesis that the time series is stationary. 

## Decomposition

```{r}
# Decompose time series into trend, seasonality, and noise 
PM2.5_comps <- decompose(PM2.5)
plot(PM2.5_comps)
```

The ADF test suggested the series is stationary; however, the decomposition reveals an upward trend and annual seasonality. 

## Autocorrelation 

```{r}
# ACF and PACF plots (lag=365 selected to represent a year)
acf2(PM2.5, 365.25, main='ACF & PACF')
```

The autocorrelation at lag 1 is very strong. The ACF tails off, while the PACF cuts off after lag 1, which suggests an AR(1) model. Let's try it out. 

# Modeling 

## Autoregression

```{r}
# AR(1) model 
sarima(PM2.5, p=1, d=0, q=0)
```

There are some problems with this model. The tail ends of the residuals deviate quite significantly from the normal distribution, according to the Q-Q plot. Furthermore, the p values of the Ljung-Box statistic are all under .05, which suggests autocorrelation within the residuals.

Let's try taking the log of the series first to see if that will help normalize the residuals.

```{r}
sarima(log(PM2.5), 1, 0, 0)
```

Taking the log actually made the residuals even less normal. Let's try taking the first difference. 

```{r}
sarima(PM2.5, 1, 1, 0)
```

That didn't solve the problem either. Let's try taking both the log and the first difference.

```{r}
sarima(log(PM2.5), 1, 1, 0)
```

No luck. 

Instead of trying to normalize the residuals and remove their autocorrelation, we can search for the optimal lag length using AIC, which balances minimizing model variance with minimizing model complexity. 

```{r}
VARselect(PM2.5,lag.max=50, type="const")
```

The optimal lag length is 19 based on AIC. Let's try it out. 

```{r}
sarima(PM2.5, 19, 0, 0)
```

The diagnostic plots for AR(19) show that the residuals have no significant autocorrelations, and all Ljung-Box statistics are larger than the threshold p-value. The tails of the residuals distribution do however still deviate from normality. This model has an AIC of 14075.71.

## Seasonality-Removed Autoregression

The decomposition of the time series suggests that there is yearly seasonality. We can manually examine yearly lagged PM2.5 obtained by taking a difference with 365th lagged term.

```{r}
PM2.5_yearly_lagged <- diff(PM2.5, lag=365) 
tsplot(PM2.5_yearly_lagged, main="Seasonality-Removed PM2.5", ylab="")
```

```{r}
acf2(PM2.5_yearly_lagged, max.lag=100, main="Autocorrelation of Seasonality-Removed PM2.5")
```

The ACF and PACF suggest that the the seasonality-removed PM2.5 still follows an AR(p) of some order p. Let's use AIC to find the optimal p. 

```{r}
VARselect(PM2.5_yearly_lagged, lag.max=50, type="const")
```

The AIC tells us that the seasonality-removed series follows an AR(5) instead of an AR(19). 

```{r}
sarima(PM2.5_yearly_lagged, 5, 0, 0)
```
This model ARIMA has a AIC of 12240.69, which is the best so far.

# Forecasting

```{r}
pred <- sarima.for(
  PM2.5_yearly_lagged, 
  p=5, 
  d=0, 
  q=0, 
  n.ahead=7,
  main="7 Day Forecast of Year-over-Year Delta in PM2.5"
)
```

Because the model was trained on the 365 day differenced data, a day's forecasted value represents its delta over that same day the previous year.  
